
---
title: "Lab 2: Trend Seasonality and Sample Autocorrelation"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(zoo)
library(tseries)
library(ggplot2)
```

## Questions

### Dataset "dataTempPG.csv"

1. **Features of the data**


Read the data into R using `read.csv()`.
```{r part1a}
df1 = read.csv("dataTempPG.csv")
```

Extract the "Summer" column and convert it into a time series in R.

```{r summer_and_ts}
summer_data = df1$Summer

# the data are yearly starting in 1919
summer_series = ts(data = summer_data,
                   start = c(1919),
                   frequency = 1)
```
Plot the time series with appropriate labels.
```{r p1plot}
p1data = fortify.zoo(summer_series)
p1 <- ggplot(p1data, aes(x = Index, y = summer_series)) +
  geom_line(color = "#E63946", linewidth = 0.65) +
  labs(
    title = "Summer Temperatures at Prince George, BC",
    subtitle = "Measured in homogenized daily minimum temperatures (C)",
    x = "Mean Temperatures (Celsius)",
    y = "Year"
  ) + theme_bw() +
  theme(panel.grid.minor = element_line(
    color = "grey90",
    linetype = "dashed",
    linewidth = 0.5
  ))
print(p1)
```


Comment on the main features of this series (e.g., seasonality, trend).


There doesn't appear to be much seasonality in the data. This makes sense, considering we are already isolating for a season (summer.) However, there does seem to be an increasing (positive) trend to the data. 


2. **Sample Autocorrelation**
    - Use `acf()` to create the autocorrelation function for the mean summer temperature data.
    - Comment on the behavior of the sample autocorrelation function.

3. **Smoothing**
    - Use `window()` to extract the portion of the time series between 1968 and 2008.
    - Plot this recent record of the data.
    - Use `rollmean()` with a moving average parameter of k=5 to add the smoothed series to the plot.
    - Add a legend to your plot.

### Dataset "LakeLevels.csv"

1. **Sample Autocorrelation Function**
    - Read the data into R and convert it into a time series.
    - Use `acf()` to create the sample autocorrelation function for the lake levels.
    - Comment on the behavior of this autocorrelation function.

2. **Decomposition (Additive Model)**
    - Use `decompose()` to decompose the time series into trend, seasonal component, and error (additive model).
    - Plot the trend, seasonal component, and error.

3. **Decomposition (Loess Method)**
    - Use `stl()` with `s.window` set to "periodic" for decomposition.
    - Plot the trend, seasonal component, and error using the loess method.
